---
title: "STA2453 - Course Documents"
output: html_document
---

# Course Documents

## September 15

- [Class 1 - Slides](lecture-notes-slides-2020/slides/sta2453-2020-lecture1.html)
- [Class 1 - lab](lecture-notes-slides-2020/lab1.html)
- [Class 1 - data](lecture-notes-slides-2020/raw_ed_data.csv)
- [Class 1 - partial lab solutions](lecture-notes-slides-2020/lab1-solution.html)

## September 29

- [Class 2 - Slides](lecture-notes-slides-2020/slides/sta2453-2020-lecture2.html)
- [Class 2 - lab](lecture-notes-slides-2020/lab2.html)
- [Class 2 - partial lab solutions](lecture-notes-slides-2020/lab1-solution.html)

## October 13th

- [Class 3 - Notes](lecture-notes-slides-2020/slides/sta2453-2020-lecture3.html)
- [Class 3 - lab](lecture-notes-slides-2020/lab3.html)
- [Class 3 - NES data](nes_data.csv)


# Marking Rubrics

## Data Analysis Rubric

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
datanal <- read_csv("dataanal_rubric.csv")
knitr::kable(datanal)
```

## Programming Rubric

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
prog <- read_csv("coding_rubric.csv")
knitr::kable(prog)
```



