---
title: "STA2453 - Course Documents"
output: html_document
---

# Course Documents

## September 15

- [Class 1 - Slides](lecture-notes-slides-2020/slides/sta2453-2020-lecture1.html)
- [Class 1 - lab](lecture-notes-slides-2020/lab1.html)
- [Class 1 - data](lecture-notes-slides-2020/raw_ed_data.csv)
- [Class 1 - partial lab solutions](lecture-notes-slides-2020/lab1-solution.html)

## September 29

- [Class 2 - Slides](lecture-notes-slides-2020/slides/sta2453-2020-lecture2.html)
- [Class 2 - lab](lecture-notes-slides-2020/lab2.html)
- [Class 2 - partial lab solutions](lecture-notes-slides-2020/lab2-solution.html)

## October 13th

- [Class 3 - Notes](lecture-notes-slides-2020/slides/sta2453-2020-lecture3.html)
- [Class 3 - lab](lecture-notes-slides-2020/lab3.html)
- [Class 3 - NES data](nes_data.csv)

## October 27th

- [Class 4 - Notes](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/rvest_pres.html)
- [Project #1 description](lecture-notes-slides-2020/sta2453_2020_project1.html)

Tutorials for different scraping techniques:

- [basic scraping](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/basic_html.html)
- [front end scraping](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/client_facing_rvest.html)
- [back end scraping](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/server_facing_rvest.html)
- [pdf scraping](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/pdf_scrape.html)
- [form filling scraping](lecture-notes-slides-2020/slides/sta2453-2020-lecture4/form_example.html)


## November 3rd
- [Working together](lecture-notes-slides-2020/slides/working-together.html)
- [Working together (pdf version)](lecture-notes-slides-2020/slides/working-together-pdf.pdf)
- [pull request template](lecture-notes-slides-2020/slides/pull_request_template.md)

### Useful resources

[Happy Git and Github for the R user](https://happygitwithr.com/)

## November 17th

- [Slides](lecture-notes-slides-2020/slides/building your first ML application.pdf)
- [Github repo](https://github.com/dwhdai/house_price_predictor)
- [Deployed application](https://david-dai.shinyapps.io/HomeListingPrice/)

# Marking Rubrics

## Data Analysis Rubric

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
datanal <- read_csv("dataanal_rubric.csv")
knitr::kable(datanal)
```

## Programming Rubric

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
prog <- read_csv("coding_rubric.csv")
knitr::kable(prog)
```



