---
title: "STA2453H1Y - Data Science Methods, Collaboration, and Communication"
subtitle: "Syllabus"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    toc: true
    toc_float: true
bibliography: syllab.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Instructor

[Prof. Josh Murray](https://www.chartdatascience.ca/josh-murray)

# Course Description

Data Science Methods, Collaboration, and Communication represents broadly, the skills required to work as a data scientist in industry or academia. This course will introduce students to these topics with an emphasis on collaboration and communication. Specifically, students will gain experience with the data science process including: 

- Problem Formulation
- Data collection and understanding
- Data wrangling
- Programming ([R](https://cran.r-project.org)/[Python](https://www.python.org))
- Data exploration
- Data vizualization
- Modelling
- Communication
- Reproducibility

Students will learn about these topics by working on case studies based on problems that data scientists face in industry and academic research. Many of the cases will involve data collected by an organization (e.g., organization or scientist), using published data, or scraping web pages. 

All projects will involve some type of collaboration or communication. Students are expected to be familiar with the application of basic statistical methods used for inference (e.g., general linear models), prediction (e.g., linear and logistic regression), and are comfortable with basic data analysis using a programming language such as R or Python. In class demonstrations will be conducted using the [R](https://cran.r-project.org) programming languages. Students have the option of submitting work with Python, however, they will have more limited support from me. 

Students will be expected to adopt a reproducible research workflow using tools such as [Github](https://github.com), and [RMarkdown](https://rmarkdown.rstudio.com), or [Jupyter](https://jupyter.org). 

Class time will be a mixture of informal lectures, class discussions, and student presentations.

# Evaluation

All work will be graded on a scale from 1 to 4 (sometimes with pluses and minuses) where:


------------------------------------------------------------------------
Grade value   Description
-----------   ---------------------------------------------------------
1             Work does not meet expectations.

2             Work meets expectations minimally, possibly missing some.

3             Good work; meets all or most expectations.

4             Excellent work; exceeds expectations.
------------------------------------------------------------------------


Grades will almost always be 2 or 3 (1’s and 4’s are rare). Generally speaking, a 2 is a B, a 3 is an A, and a 4 is an A+.

```{r, echo=FALSE}
sched <- read_csv(file = "sta2453grading.csv", 
                  na = character(), 
                  col_types = cols(Item = col_character(), 
                                   Description = col_character(), 
                                   Value = col_character()))
knitr::kable(sched)
```



# Course Schedule

This is a half-credit that meets in both the fall and winter terms.  Class meeting will occur approximately bi-weekly.

```{r, echo=FALSE}
library(kableExtra)
library(tidyverse)

sched <- read_csv(file = "sta2453_2020schedule.csv", 
                  na = character(), 
                  col_types = cols(Class = col_character(), 
                                   Date = col_character(), 
                                   Description = col_character(), 
                                   Reading = col_character(), 
                                   Due = col_character()))
knitr::kable(sched) %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(3, bold = T) %>%
  row_spec(c(1,3,5,8,10,12,13,19,20,22,23,25,27,29), bold = T, color = "white", background = "#3bd432")
```

## Required Readings

[ISLR](http://faculty.marshall.usc.edu/gareth-james/ISL/) (@james2013introduction) and [CASI](https://web.stanford.edu/~hastie/CASI/) (@efron2016computer) are both are freely available.

Other readings may be assigned during the course.

# Course Books


